### 2017-08-28 marking_response/response.tex

\noindent
\textcolor{blue}{
	\textit{Given this thesis is focused on a known environmental influence on cancer (obesity), it was a surprise the section on environmental factors associated with cancer incidence (1.2.3) is underdeveloped and superficial.
	}
}\\

\noindent
I understand the Examiner's concern about the relatively brief description of the known environmental factors that affect cancer incidence.
As you may know, the focus of this project was the relationship between cancer and obesity, and how obesity affects cancer (which is covered more in depth in Section 1.3.2).

Section 1.2 was written as a section introducing cancer in general, and was not specifically focused on any single cancer type, nor environmental factor.
So it follows that Section 1.2.3 describes, in very general terms, how \textit{some} (not \textit{all}) environmental factors can influence the cancer incidence, and I think the amount of information placed in this subsection is sufficient for this purpose.
Additionally, the environmental factors chosen to be included in this section was based on either (1) (very) strong evidence of their role in cancer onset/progression in the literature (in terms of both epidemiological and biological evidence/association), or (2) some sort of relationship with obesity.

Perhaps, on request, I could emphasise these points within Section 1.2.3 and guide the readers to Section 1.3.2 for specific details on the effect of obesity on cancer.
\\

@@

\noindent
As mentioned above, it was of interest to find novel obesity-associated pathways or genes that were common across multiple cancer types, so DEG analysis was carried out on all available cancer types, regardless of the tissue of origin.
The purpose of the simulation was to make sure, statistically, that the number of DEGs identified between multiple cancer types were significant, and not by chance.
However, as you may know from the report, majority of these DEGs would have been false positives.

Again, using only the cancer types that were known to be associated with obesity may have aided the discovery of significant pathways and/or genes.
\\

\noindent
\textcolor{blue}{
\textit{On page 110, it is suggested that as the obesity-associated genetic signatures were derived from breast cancer datasets, that it made sense that these signatures were significant in the bladder cancer data sets as `basal-like' and `luminal-like' subtypes had been identified in bladder cancer.
As adenocarcinomas (epithelial-cell derived, and glandular in origin) all have these characteristics due to cell of origin context, I am not convinced by interpretations made in section 5.1.2, and feel the statements made are ungrounded.
}
}\\

\noindent
I would like to point out that Damrauer \textit{et al.} not only identified luminal and basal-like subtypes in bladder cancer, but they also showed that the molecular signatures (both gene expression pattern and gene set enrichment) between the breast and bladder cancer subtypes had striking similarity.
This suggests that the underlying biology of the subtypes between the two cancer types are very similar.
I am not aware of any other cancer type showing this sort of similarity.

It could be possible that many other adenocarcinoma may have these characteristics due to their cell of origin context.
However, based  on my very limited knowledge of the broad field of cancer biology, breast cancer-like subtypes  have not yet been identified in any other cancer types, let alone compared with the breast cancer subtypes like in the Damrauer \textit{et al.} paper.

Based on the current knowledge of adenocarcinoma subtypes and the unexpected analysis results from this thesis, I personally think that the interpretation made on page 110  was the best explanation one could make out from these evidence.
\\

@@

\newpage

\section*{Response to the comments made by Examiner ``1137''}
\label{sec:response_to_comments_from_examiner_1137}

\noindent
\textcolor{blue}{
	\textit{Individually published papers could have been critically appraised in the Introduction when it was introduced.
	}
}\\

\noindent
%TODO: "critically appraise" the three papers.
Creighton \textit{et al.}, Fuentes-Mattei \textit{et al.} and Gatza \textit{et al.} papers have been more thoroughly introduced in Section 1.4.
\\

@@

\noindent
\textcolor{blue}{
	\textit{The effect of obesity on treatment response in cancer, and the influence of obesity on the likelihood of initial diagnosis, may also have implications for obesity-gene expression associations found in the literature.
	}
}\\

\noindent
\\

\noindent
\textcolor{blue}{
	\textit{Potential effects on gene expression in tumours by lifestyle factors associated with obesity, but not recorded in most data sets was not discussed in great detail, however this is relevant to the concepts explored later in Section 3.3.
	}
}\\

\noindent
Adjusting the linear models with other clincial variables such as smoking would have been interesting to see.
Extra paragraph has been added to Section 1.2.3 and a sentence has also been added towards the end of Section 5.1.1 to address this point.
\\

@@

\noindent
\\

\noindent
\textcolor{blue}{
	\textit{The interesting literature from \textit{in vitro} studies about the effect on cancer cells of factors produced by adipocytes in culture was not explored.
	}
}\\

\noindent
\\

\noindent
\textcolor{blue}{
	\textit{Potential mechanisms of obesity-altering hormone receptor expression could have been explored in more depth \textbf{at the level of cancer cells}.
	The effects of endocrine factors like leptin and the IGFs were discussed in general but not the cancer cell type-specific gene expression changes they cause.
	}
}\\

@@

\noindent
\textcolor{blue}{
	\textit{The justification presented about the appropriateness of using SVD and transformation matrices is convincing but I wonder about the effects of small numbers of probe sets/RNAs with specific biological connotations dominating some metagenes.
	}
}\\

\noindent
I agree that there will be some probe sets/RNAs that would have influenced the metagene scores more than the others, and the identification of these influencial probes/RNAs would be of great interest.
However, I feel like an investigation  of the most informative probes/RNAs for each of the obesity- or pathway-associated signatures (or any genetic signature in question) would require an effort of a whole new project.
\\

\noindent
\textcolor{blue}{
	\textit{Cancer patients can have vast differences in BMI at different points in their treatment.
	Was the clinical timing of weight measures standardised in some way to reduce the impact of this problem, e.g. using weight at initial diagnosis?
	}
}\\

\noindent
This is a good point, and only Creighton \textit{et al.} have specifically stated in their paper that the patient BMI was measured at the time of diagnosis.
Fuentes-Mattei \textit{et al.} paper did not describe when the patient BMI was measured.
Although I am not aware of how and when clinicians record patient data, I would suspect they would have measured BMI at the time of diagnosis to keep the data consistent with other variables (e.g. tumour grade).
\\

@@

\noindent
This will be considered after discussion with Mik and HOD.
\\

\noindent
\textcolor{blue}{
	\textit{Add additional description about what basic quality control was possible for each data set (perhaps in section 2.3), and the results of the QC.
	The discussion in section 5.2.2 is hard to interpret without this.
	}
}\\

\noindent
\\

\noindent
\textcolor{blue}{
	\textit{It would be interesting to explore whether the differences in risk ratio between cancer types was due to technical issues, or represented real biology.
	Discuss whether the biology underpinning differences between cancer types in terms of obesity risk ratios in Table 1 could be explored statistically using genomic data.
	}
}\\

\noindent
Though there would be some technical issues in the individual studies include in the meta-analysis study by Renehan \textit{et al.}, I suspect that there is an effect of obesity on certain types of cancer.
I am not sure whether there are any genomic data that could clarify the Examiner's question, as it is possible that obesity affects many different biological functions/pathways of a cell.
However, I wonder if it would be possible to use SVD, or some other weighting scores, that take these risk ratios into account and essentially ``normalise'' the genetic signals (e.g. gene expression and/or methylation) across different cancer types, and thereby focus on the underlying biology of each cancer type.
\\

### 2017-08-28 thesis/discussion/discussion.tex
tmp commit

It was possible that all of these results were due to the sample quality (and hence due to the quality of the data), or perhaps many of the samples in these data sets had breast cancers due to a cause unrelated to their obesity status.
As many epidemiological studies have pointed out, obesity is a strong risk factor for cancer (covered in \cref{sec:obesity_and_cancer}), but this does not necessarily mean that all cancer in obese patients develop cancer as a result of obesity.
In other words, the cancer identified in obese patients may have been caused by another, completely different biological mechanism.
Thus the tumour samples taken from obese and overweight patients will be a mixture of samples that are driven by obesity-related mechanisms and mechanisms that are not related to obesity.
These different subgroups of tumours (obesity-driven or not) cannot be easily identified and isolated.
This may affect the detection of the truly obesity-associated genes, as some of the tumour samples from the patients will have nothing to do with obesity and cancer, and hence misclassified into wrong groups.

@@

In this project, the lack of association of the obesity-associated genetic signatures with any of the cancer types could have been due to the quality of both the raw data sets and the genetic signatures used in this project.
Furthermore, the use of \gls{bmi} may have been insufficient to describe the patients' obesity status, and would likely require more informative clinical measurements, such as waist circumference, to aid the classification.
Therefore, careful selection of the raw data sets with sufficient, and if possible, additional information (such as waist circumference) about the patients and well-validated genetic signatures will be essential for future investigations.

>>>>

It was possible that all of these results were due to the sample quality (and hence due to the quality of the data), or perhaps many of the samples in these data sets had breast cancers due to a cause unrelated to their obesity status.
As many epidemiological studies have pointed out, obesity is a strong risk factor for cancer (covered in \cref{sec:obesity_and_cancer}), but this does not necessarily mean that all cancer in obese patients develop cancer as a result of obesity.
As briefly mentioned in \cref{ssub:environmental_factors}, there are many other environmental factors that can cause cancer and patients in these studies may have been exposed to these factors, but not recorded or measured the exposure in the clinical data for adjustment in the analysis.

In other words, the cancer identified in obese patients may have been caused by another, completely different biological mechanism.
Thus the tumour samples taken from obese and overweight patients will be a mixture of samples that are driven by obesity-related mechanisms and mechanisms that are not related to obesity.
These different subgroups of tumours (obesity-driven or not) cannot be easily identified and isolated.
This may affect the detection of the truly obesity-associated genes, as some of the tumour samples from the patients will have nothing to do with obesity and cancer, and hence misclassified into wrong groups.

@@

In this project, the lack of association of the obesity-associated genetic signatures with any of the cancer types could have been due to the quality of both the raw data sets and the genetic signatures used in this project.
Furthermore, the use of \gls{bmi} may have been insufficient to describe the patients' obesity status, and would likely require more informative clinical measurements (such as waist circumference) to aid the classification.
Therefore, careful selection of the raw data sets with sufficient, and if possible, additional information (such as waist circumference) about the patients and well-validated genetic signatures will be essential for future investigations.

### 2017-08-28 thesis/intro/intro.tex
tmp commit

Diet and obesity are also known risk factors for many types of cancers \citep{Ames1995,Calle2004}.
Epidemiological evidence of obesity being a risk factor for many cancers has been well established, yet the mechanisms of its contribution to cancer is still  speculative \citep{Calle2003,Kelesidis2006}.
Although there are a lot of possible mechanisms, there is growing evidence of the role of chronic inflammation caused by increased fat mass in the body \citep{Kelesidis2006,Lumeng2011,Hernandez2013}.
Free fatty acids from excess adipose tissue and/or diet can be recognised by \gls{tlr} (specifically \gls{tlr}4), which can then activate the \gls{nfkb} pathway, leading to the activation of one of the Hallmarks of Cancer (tumour-promoting inflammation) \citep{Lumeng2011}.
There are many other environmental factors associated with cancers, thus limiting the exposure to detrimental environmental factors is crucial for the prevention of many cancers.

>>>>

Diet and obesity are also known risk factors for many types of cancers \citep{Ames1995,Calle2004}.
Epidemiological evidence of obesity being a risk factor for many cancers has been well established, yet the mechanisms of its contribution to cancer is still  speculative \citep{Calle2003,Kelesidis2006}.
Although there are a lot of possible mechanisms, there is growing evidence of the role of chronic inflammation caused by increased fat mass in the body \citep{Kelesidis2006,Lumeng2011,Hernandez2013}.
Free fatty acids from excess adipose tissue and/or diet can be recognised by \gls{tlr} (specifically \gls{tlr}4), which can then activate the \gls{nfkb} pathway, leading to the activation of one of the Hallmarks of Cancer (tumour-promoting inflammation) \citep{Lumeng2011}.

In addition to the mechanisms described above, there is a growing body of evidence where some environmental factors may affect the transcriptional regulation of a set of genes in certain cancer type.
A relatively recent review by \citet{Perez-Solis2016} describes potential mechanism in which obesity, alcoholism and smoking may regulate gene expression in breast cancer cells.
By affecting the so-called ``Master'' transcription factors, environmental factors are able to affect the expression of the genes that are downstream to the Master transcription factor, and as a result, lead to the development of cancerous cells \citep{Perez-Solis2016}.
It is not surprising that different environmental factors affect different Master transcriptional factor, and therefore have different biological mechanism of gene regulation.

It is also worth noting here that multiple  environmental factors may be exposed to an individual at any given point of time, and perhaps a combination of multiple environmental factors may have more significant effect on the individual.
There are many other environmental factors associated with cancers that have not been mentioned here.
Nevertheless, it is evident that limiting the exposure to detrimental environmental factors is crucial for the prevention of many cancers.
